Predator clusters

### 2023-02-23 reports/04_pred_clus.qmd
Indicator species and predator cluster maps

---
title: "Predator clusters"
format:
  html:
    code-fold: true
editor: visual
output-dir: "docs"
---

@@

tar_load(c("predators_stations", "predators_abundant", "sightings_clust",
           "zoop"), 
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
```

@@

total_stations <- n_distinct(zoop$amlr.station)
pred_frac <- predators_stations %>%
  group_by(amlr.station, species) %>% 
  summarize(count = sum(count), .groups = "drop") %>%
  group_by(species) %>%
  summarize(stations_present = n(),
            station_frac = stations_present / total_stations,
            .groups = "drop") %>%
  arrange(desc(station_frac))
cutoff_row <- sum(pred_frac$station_frac >= 0.05)

@@

Hierarchical clustering of stations by predator community. Chose 5 clusters a priori to match Dietrich et al. (2021). Note: `r nrow(zoop)` stations were included in the zooplankton clustering, but only `r length(sightings_clust$labels)` stations were used for predator clustering.

@@

# Join with zooplankton clusters
both_clust <- zoop %>% 
  transmute(amlr.station, 
            zoop_clust = factor(`Winter Cluster factor`)) %>% 
  right_join(tibble(amlr.station = names(clust),
                    pred_clust = factor(clust)),
             by = "amlr.station")

@@

The above mosaic plot shows the distribution of stations by predator and zooplankton stations. For example, the bottom left rectangle indicates 9 stations were assigned to zooplankton cluster 1 and predator cluster 1. Rectangles color-coded by zooplankton cluster using the color palette from Dietrich et al. 2021 Fig 3a.

@@

>>>>

---
title: "Predator clusters"
format:
  html:
    code-fold: true
    toc: true
editor: visual
output-dir: "docs"
---

@@

tar_load(c("predators_stations", "predators_abundant", "sightings_clust",
           "zoop_sf"), 
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
```

## Predator clusters

@@

total_stations <- n_distinct(zoop_sf$amlr.station)
pred_frac <- predators_stations %>%
  group_by(amlr.station, species) %>% 
  summarize(count = sum(count), .groups = "drop") %>%
  group_by(species) %>%
  summarize(stations_present = n(),
            station_frac = stations_present / total_stations,
            .groups = "drop") %>%
  arrange(desc(station_frac))
cutoff_row <- sum(pred_frac$station_frac >= 0.05)

@@

Hierarchical clustering of stations by predator community. Chose 5 clusters a priori to match Dietrich et al. (2021). Note: `r nrow(zoop_sf)` stations were included in the zooplankton clustering, but only `r length(sightings_clust$labels)` stations were used for predator clustering.

@@

# Join with zooplankton clusters
both_clust <- zoop_sf %>% 
  transmute(amlr.station, 
            zoop_clust = factor(`Winter Cluster factor`)) %>% 
  right_join(tibble(amlr.station = names(clust),
                    pred_clust = factor(clust)),
             by = "amlr.station")

@@

The above mosaic plot shows the distribution of stations by predator and zooplankton clusters. For example, the bottom left rectangle indicates 9 stations were assigned to zooplankton cluster 1 and predator cluster 1. Rectangles color-coded by zooplankton cluster using the color palette from Dietrich et al. 2021 Fig 3a.

@@

## Indicator species

Which species best represent each cluster? Using Dufrene-Legendre indicator species analysis. Apply indicator value cutoff at 25%.

```{r}
#| label: indicator_species

species_wide <- pivot_wider(predators_abundant,
                            names_from = species,
                            values_from = count,
                            values_fill = 0) %>% 
  column_to_rownames("amlr.station")
cluster_indval <- labdsv::indval(species_wide, clust)
indval_mtx <- cluster_indval$indval
indval_fmt <- apply(indval_mtx, 2, function(x) {
  ord <- order(x, decreasing = TRUE)
  result <- ifelse(x >= 0.25, 
         str_glue("{rownames(indval_mtx)} ({round(x * 100)}%)"),
         NA) %>% 
    `[`(ord) %>% 
    na.omit() %>% 
    paste(collapse = "<br>")
  ifelse(result == "", "None", result)
})
as_tibble(t(indval_fmt)) %>% 
  knitr::kable(format = "html", escape = FALSE) %>% 
  kableExtra::kable_styling()

According to the 25% cutoff, clusters 1 and 4 lack indicator species. In agreement, every species' highest indicator value falls in clusters 2, 3, or 5. This suggests to me 5 clusters is a poor choice for the predators and we should use fewer. I'll calculate the gap statistic for the hierarchical cluster analysis to quantitatively choose a data-informed number of clusters and repeat this entire analysis. For now, continuing with 5 clusters.

I don't know enough about the Antarctic Peninsula megafauna to say whether the indicator species cluster together logically. The fulmarine/*Daption* petrels are split across Clusters 2 and 3. Cluster 2 lumps Cape Petrel and Antarctic Fulmar with the prion-like Blue Petrel. Cluster 3 combines the fulmarine Antarctic and Snow Petrels with the Kerguelen Petrel and fur seals. Cluster 5 seems to have the most ice-associated species (phocids and Adelie penguins).

## Cluster maps

What geospatial patterns do we see in the predator clusters? Cluster 5 is mostly found in the Bransfield Strait, which makes sense for the Adelie Penguin-crabeater seal-leopard seal complex. Cluster 2 (Antarctic Fulmar, Cape Petrel, and Blue Petrel) tends to show up oceanward of the South Shetland Islands. Cluster 3 (fur seal, snow petrel, Kerguelen petrel, and Antarctic Petrel) seems the most variable; found both inside and outside the Strait depending on the year.

```{r}
#| label: cluster_maps

pred_clust_sf <- tibble(amlr.station = names(clust),
                        pred_clust = factor(clust)) %>% 
  left_join(select(zoop_sf,
                   amlr.station,
                   geometry,
                   Year),
            by = "amlr.station") %>% 
  st_as_sf() %>% 
  st_transform(ant_proj())

map_lim <- sf::st_bbox(pred_clust_sf) %>% 
  project_bbox() %>% 
  expand_bbox(factor = 1.2)

ant_basemap(map_lim) +
  geom_sf(aes(color = pred_clust), pred_clust_sf, size = 2, shape = 15) +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = c(-60, -55)) +
  scale_y_continuous(breaks = c(-63, -62, -61, -60)) +
  coord_ant(map_lim) +
  theme(legend.position = "bottom")
```
